---
title: "Chapter 10 examples"
author: "thm"

format: 
  html:
    theme: custom.scss
    embed-resources: true
    code-tools: true

csl: ecology.csl

editor: visual
bibliography: references.bib
---

```{r}
#| label: setup
#| message: false
#| echo: "fenced"

knitr::opts_chunk$set(
  echo = "fenced"
  )

library(emmeans)
library(lattice)
library(redres)   # For interactive (shiny) diagnostic plots of lmer models
library(lmerTest)  
library(car)       
library(latex2exp)
library(kableExtra)
library(ggfortify)
library(broom)
library(tidyverse)

options(knitr.kable.NA = "")

# Installs redres, just delete the "#" 
# devtools::install_github("goodekat/redres")
```

## Box 10.1\| Mites on leaves

@Walter.ODowd1992 examined the role of domatia (small shelters produced by plants for arthropod symbionts) in determining the numbers of mites on leaves of plant species with domatia. They did an experiment using 14 pairs of leaves (randomly chosen) with one leaf in each pair with shaved domatia and the other as a control (normal domatia). The response variable was total number of mites per leaf, which @Walter.ODowd1992 transformed to $ln(0.5 + (mite~\times~10))$, ostensibly to improve normality and homogeneity of variances among treatments, the 0.5 added because of zeros although multiplication by ten seemed unnecessary. The pairs of leaves were considered blocks. The data were analyzed with the blocks as fixed (model 1).

$$
y_{ij} = \mu + \alpha_i + \beta_j + \epsilon_{ij}
$$ {#eq-Fixed}

```{r}
#| label: tbl-dataImport
#| tbl-cap: "First six observations of the data provided by @Walter.ODowd1992."

mites <- read_csv(
  "walter.csv",
  col_select = c(1:4), # to avoid the columns the textbook authors calculated
  show_col_types = FALSE
  )
colnames(mites) <- tolower(colnames(mites))

mites$treat <- factor(
  mites$treat,
  levels = c(1,2),
  labels = c(
    "Control",
    "Shaved"
    )
  )

mites$block <- factor(mites$block)

kbl(
  head(mites),
  col.names = c(
    "Leaves",
    "Pair",
    "Treatment",
    "Number of Mites"
    )
  ) |>
  kable_classic(
    full = FALSE
    )
```

```{r}
#| label: tbl-fixedANOVA
#| tbl-cap: "ANOVA summary for the data provided by @Walter.ODowd1992 and presented in Box 10.1 of the textbook."
#| fig-show: hide

modFixed <- aov(log(0.5 + mite*10) ~ treat + block, data = mites)

plot(modFixed) # assumption check

kbl(
  Anova(modFixed),
  digits = c(3,0,3,3)
  ) |>
  kable_classic()
```

Given that the pairs were chosen at random, it seems that the blocks should have been treated as a random factor (model 2).

$$
y_{ij} = \mu + \alpha_i + B_j + \epsilon_{ij}
$$ {#eq-random}

```{r}
#| label: tbl-mixedANOVA
#| tbl-cap: "Linear mixed-effect model results for the data provided by @Walter.ODowd1992."
#| tbl-subcap: 
#|     - "Fixed effect."
#|     - "Random effect."
#| layout-nrows: 2

modMixed <- lmer(log(0.5 + mite*10) ~ treat + (1|block), data = mites)

# launch_redres(modMixed)

outMixed <- anova(modMixed)

kbl(
  outMixed,
  col.names = c(
    "SS",
    "MS",
    "df1",
    "df2",
    "F",
    "P"
    ),
  digits = c(3,3,0,0,3,4)
  ) |>
  kable_classic()

kbl(
  rand(modMixed),
  col.names = c(
    "Parameters",
    "log-likelihood",
    "AIC",
    "LRT",
    "df",
    "P"
    ),
  digits = c(0,3,3,3,0,4)
  )
```

In this case, the qualitative results for treating blocks as fixed or random were the same (@tbl-fixedANOVA, @tbl-mixedANOVA). Removing domatia resulted in few mites.

## 
